### 2018-03-08 Writing/disparity_review.tex
Edits to methods

\section{Use appropriate methods} 
In the vast majority of morphological disparity analysis, the methodological approach ought to be multidimensional.
Multidimensional analysis can be a really powerful tool to analysis complex datasets but also comes with many mathematical caveats and limitations.
As any great power, it comes with great responsibilities.

One first step in any statistical analysis is to visualise the data to understand its distribution and limitations at a glance.
Unfortunately, our brains and physical world only allows us to understand a maximum of three dimensions at the same time.
Therefore one of the main challenges is multidimensional analysis is how to reduce the number of dimensions whether this is for analysis (see box \ref{box_ordination}) or visualisation.

\subsection{Visualising the multidimensional problem}
\label{visualisation}
Because of the limitation of our physical world, it is important to remember first and foremost that visual interpretation of multidimensional data can easily be misleading.
For example, it is incorrect to assume that because some groups overlap on a bivariate (or 3D) plot, they overlap in the whole multidimensional space (especially if the space is not euclidean!).
Note that conversely, when two groups don't overlap on a two or three dimensional plot, they \textit{do not} overlap in the multidimensional space.
Furthermore, multidimensional spaces are not necessary Euclidean.
Such spaces can have some non-intuitive properties, for example straight lines in some dimensions are not straight in the space or, less intuitively, distance between two points A and B might not be equal to the distance between B and A [CITE Sylvain].

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/dimensionsOverlap.pdf}
\caption{\small{Two groups overlap in some dimensions do not necessary overlap in all the dimensions. Conversely, two groups not overlapping on any dimensions don't overlap in all dimensions. In this example, no group overlaps in the multidimensional space despite the appearances.}}
\label{Fig:RF_results_best}
\end{figure}

It is thus always important to be entirely clear about what is represented in multidimensional spaces visualisation.
In fact, it can be easy (intentionally or not) to mislead a reader into over/under interpreting a spatial pattern if not all the key properties of the space are present in the graph.
We believe that these should always include the following:
(1) the percentage of the variation of each axis of the plot (i.e. is the majority - or all - the variation of the data present in the plot or only a fraction?);
(2) the scale of the axis (ideally, if the space is orthogonal, it is important to keep the same scale between the axis);
and (3) the error associated with each point, whether this is in term of measurement error (on ordination plots for example) or in standard variation (in further analysis, e.g. disparity through time).
Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worth, it won't be what it seems.

\subsection{[Box]To ordinate or not to ordinate (that is the multidimensional question)?}
\label{box_ordination}
One of the most common approach to analyse multidimensional data is to use ordination techniques.
This consists in transforming the dataset along its major axis of variation through various mathematical methods [Legender \& Legendre].
This can be done through three main analysis pipelines:
\begin{enumerate}
    \item data -> distance matrix -> ordination
    \item data -> zscores -> PCA
    \item data -> Procrustes superimposition -> projection into tangent space -> PCA
\end{enumerate}
%TG: Check with Melanie for the figure? - apparently she has one.

%Why to ordinate:
These pipelines allow to reduce the dimensionality of the data.
This is advantageous for plotting and visualising the data (see Visualisation section \ref{visualisation}) which can reveal properties of the morphospace not captured by the metrics (see Metrics section \ref{metrics}).
Additionally, ordinating the data can allow to further reduce the number of dimensions by only selecting a number of axis based on the hypothesis (e.g. selecting only the axis that describe up to 95\% of the variation).
This can be done through the scree plots (cuts off, majority axis, etc.) or other statistical analysis (Horns parallel, etc.).
In the case of geometric morphometric data (point 3 above), ordination can be particularly relevant since it conserves mathematical properties of the data while efficiently reducing the dimensions: the PCA reflects the number of degrees of freedom (resulting from the superimposition), its geometry is conserved (coordinates remain Euclidean when projected into tangent space and ordinated), etc.
This has clear advantages for interpreting the results in a lower number of dimensions.
For example, the axis will actually represent gradients of biological variation (e.g. elongation and flattening of the beak).
This also allows easy interpretation of the proximity of the elements in space which reflects shared biological characters.

However, some caveats can rise from such transformation of the data.
For example, in the case of an ordination in a geometric morphometric data context, not using all the axes from the ordination (e.g. only the few first) can lead to misinterpretation of the results erroneous statistical results.
Furthermore, interpreting the biological variation along the axis is always a \textit{post-hoc} procedure and might bear little meaning regarding the disparity question (e.g. the elongation and flattening of the beak when the question is about flight pattern disparity).

In fact, in many cases, depending on the question, ordination might actually not be necessary.
For example, if the disparity metric relies on all the data (for example the sum of variance or the sum of ranges) it is not necessary to calculate it on ordinated data [e.g. CITE Close].
Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
For example, when the analysed data is non Euclidean (e.g. discrete morphological characters), the ordination can be problematic (character hierarchy - inapplicable -, polymorphisms, missing data), and the resulting non-Euclidean ordinated space can be difficult to interpret.
This can be problematic when comparing the position of groups in the multidimensional space since the measured distances might not be comparable among each other.
Furthermore, the \textit{post-hoc} interpretation of the gradient of variation on the ordination axis can be simply impossible or meaningless biologically (e.g. the ratio of data coded as 1 rather than 0).

Regarding these advantages and inconvenient of ordination technique, we strongly suggest to not automatically ordinate data in multidimensional analysis and to carefully consider whether the question can be answered without ordination.
\subsection{[Box] end}

\subsection{Phylogeny}
For many disparity analysis looking at macroevolutionary questions, phylogeny is often taken into account, whether for including estimates of ancestors or to add a phylogenetic structure to the data.
For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE Adams] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
and will thus not covered further in this review.

\subsubsection{Ancestral states estimation}
When looking at disparity in the past (e.g. disparity-through-time) ancestral states estimations are a common approach.
Ancestral states estimations can be used in two ways:
\begin{enumerate}
\item pre-ordination: the estimations is done before transformation of the data (e.g. ordination, distance matrix, etc.) and is simply based on the original dataset[CITE].
\item post-ordination: the estimation is done else after transformation of the data by estimating the ancestral states based on the transformed matrix (e.g. the ordination scores) [CITE].
\end{enumerate}
Pre-ordination ancestral states estimation can change the ordinated space geometry (i.e. the relation between the points not estimated) and implies longer computational method but once the morphospace is defined, it will remain constant (i.e. its properties won't change).
Post ordination ancestral states estimation will not change the ordinated space geometry and will be faster to compute, however, it will change the morphopspace's properties such as the number of parameters (i.e. elements in the space - this can be problematic for statistical tests down the line) or the variance per axis [CITE Graeme's paper to be?].

>>>>

\section{Use appropriate methods} 
In the vast majority of morphological disparity analyses, the methodological approach will be multidimensional.
Multidimensional analysis can be a powerful tool for analysing complex datasets but also comes with many mathematical caveats and limitations.
As with any great power, it comes with great responsibility. % NC: Ha are we citing spiderman too? Awesome
The first step in any statistical analysis should be to visualise the data to understand its distribution and limitations.
Unfortunately, we can only perceive a maximum of three dimensions at the same time.
Therefore one of the main challenges in multidimensional analysis is how to reduce the number of dimensions, both for visualisation and analysis (see box \ref{box_ordination}).

\subsection{Visualising the multidimensional problem}
\label{visualisation}
It is important to remember that visual interpretations of multidimensional data can easily be misleading.
For example, it is incorrect to assume that because some groups overlap on a bivariate or three dimensional plot, that they overlap across the whole multidimensional space (especially if the space is not Euclidean).
Conversely, when two groups do not overlap on a two or three dimensional plot, this does not mean they never overlap in the multidimensional space.
Furthermore, multidimensional spaces are not necessary Euclidean.
Such non-Euclidean spaces often have non-intuitive properties, for example straight lines in some dimensions are not straight in the space and, even less intuitively, the distance between points A and B might not be equal to the distance between points B and A [CITE Sylvain].

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/dimensionsOverlap.pdf}
\caption{\small{Two groups that overlap in some dimensions do not necessary overlap in all dimensions. 
Conversely, two groups that do not overlap in some dimensions can overlap in other dimensions. 
In this example, no group overlaps in the multidimensional space despite the appearances.}} 
% NC: Might need to explain more clearly, I'm not 100% sure I got your meaning
\label{Fig:RF_results_best}
\end{figure}

It is therefore vital to be clear about what is represented in multidimensional space visualisations.
In fact, it can be easy (intentionally or not) to mislead a reader into over/under interpreting a pattern if all the key properties of the space are not present in the graph.
We suggest visualisations should include the following:
(1) the percentage of the variation explained by each axis of the plot (i.e. is the majority - or all - the variation of the data present in the plot or only a fraction?);
(2) the scale of the axes (ideally, if the space is orthogonal, it is important to keep the same scale across the axes);
and (3) the error associated with each point, whether this is in terms of measurement error (on ordination plots for example) or some other measure of standard variation (in further analysis, e.g. disparity through time).
Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worst, it will be misleading.

\subsection{[Box]To ordinate or not to ordinate (that is the multidimensional question)?}
\label{box_ordination}
One of the most common approaches to analysing multidimensional data is to use ordination techniques.
These consist of transforming the dataset along its major axis of variation through various mathematical methods [Legender \& Legendre].
This is generally done via three analysis pipelines:
\begin{enumerate}
    \item data -> distance matrix -> ordination
    \item data -> zscores -> principal components analysis (PCA)
    \item data -> Procrustes superimposition -> projection into tangent space -> PCA
\end{enumerate}
%TG: Check with Melanie for the figure? - apparently she has one.

%Why to ordinate: % NC: This box seems quite long, can it be reduced?
These pipelines reduce the dimensionality of the data.
This is advantageous for plotting and visualising the data (see Visualisation section \ref{visualisation}) which can reveal properties of the morphospace not captured by the metrics (see Metrics section \ref{metrics}).
Additionally, after ordinating the data it is possible to further reduce the number of dimensions by only selecting a certain number of axes (e.g. selecting only the axes that describe up to 95\% of the variation).
In the case of geometric morphometric data (point 3 above), ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions: the PCA reflects the number of degrees of freedom (resulting from the superimposition), its geometry is conserved (coordinates remain Euclidean when projected into tangent space and ordinated), etc. % NC: is the stuff after the : needed in this sentence?
This has clear advantages for interpreting the results.
For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak REF).
This also allows easy interpretation of the proximity of the elements in space which reflect shared biological characters.
%%%%%%%%%%%
However, some caveats can arise from such transformations of the data.
For example, in the case of an ordination in a geometric morphometric data context, not using all the axes from the ordination (e.g. only the first few) can lead to misinterpretation of the results.
Furthermore, interpreting biological variation along the axes is always a \textit{post-hoc} procedure and might have little relation to the disparity question (e.g. if the first few axes represent the elongation and flattening of the beak, but the question is about wing disparity).
In many cases, depending on the question, ordination might not be necessary.
For example, if the disparity metric being used (see below) relies on all the data (for example the sum of variance or the sum of ranges) it is not necessary to calculate it on ordinated data [e.g. CITE Close].
Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
For example, when the analysed data is non-Euclidean (e.g. discrete morphological characters), the ordination can be problematic (due to character hierarchy, inapplicable characters, polymorphisms, and missing data), and the resulting non-Euclidean ordinated space can be difficult to interpret.
This can be problematic when comparing the position of groups in the multidimensional space, as the measured distances might not be comparable.
Furthermore, the \textit{post-hoc} interpretation of the gradient of variation on the ordination axes can be impossible or biologically meaningless (e.g. the ratio of data coded as 1 rather than 0). % NC: Need to explain this point more clearly

Because of these points, we strongly recommend that data for multidimensional analyses is not automatically ordinated, and that careful consideration is given to whether the question can be answered without ordination.
\subsection{[Box] end}

\subsection{Phylogeny}
For many disparity analyses of macroevolutionary questions, phylogeny is often taken into account, whether for including estimates of ancestors or to add phylogenetic structure to the data.
For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE Adams] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
and therefore will not covered further in this review.

\subsubsection{Ancestral state estimation}
When looking at disparity in the past (e.g. disparity-through-time) ancestral state estimations are a common approach.
Ancestral state estimation can be performed in two ways:
\begin{enumerate}
\item pre-ordination: the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original dataset[CITE].
\item post-ordination: the estimation is done after transformation of the data by estimating the ancestral states based on the transformed matrix (e.g. the ordination scores) [CITE].
\end{enumerate}
Pre-ordination ancestral state estimation can change the ordinated space's geometry (i.e. the relationship between the points not estimated) and implies longer computational method but once the morphospace is defined, it will remain constant (i.e. its properties won't change).
Post ordination ancestral states estimation will not change the ordinated space geometry and will be faster to compute, however, it will change the morphopspace's properties such as the number of parameters (i.e. elements in the space - this can be problematic for statistical tests down the line) or the variance per axis [CITE Graeme's paper to be?].
